My Publication Title --- Single Author

### 2025-06-28 paper/quarto_aaai/paper.tex
Adjust methods

In this section we propose our novel counterfactual training objective.
In CT, we combine ideas from adversarial training, counterfactual
explanations, and energy-based modelling with the explicit goal of
aligning representations with plausible explanations that comply with
user-defined actionability constraints.

In the context of counterfactual explanations, plausibility has broadly
been defined as the degree to which counterfactuals comply with the
underlying data-generating process
\citep{altmeyer2024faithful, guidotti2022counterfactual, poyiadzi2020face}.
Plausibility is a necessary but insufficient condition for using CEs to
provide algorithmic recourse (AR) to individuals (negatively) affected
by opaque models. To be actionable, AR recommendations must also be
attainable. A plausible CE for a rejected 20-year-old loan applicant,
for example, might reveal that their application would have been
accepted, if only they were 20 years older. Ignoring all other features,
this would comply with the definition of plausibility if 40-year-old
individuals were in fact more credit-worthy on average than young
adults. But of course this CE does not qualify for providing actionable
recourse to the applicant since \emph{age} is not a (directly) mutable
feature. Counterfactual training aims to improve model explainability by
aligning models with counterfactuals that meet both desiderata:
plausibility and actionability. Formally, we define explainability as
follows:

\begin{definition}[Model Explainability]
\label{def-explainability}
Let $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denote a supervised classification model that maps from the $D$-dimensional input space $\mathcal{X}$ to representations $\phi(\mathbf{x};\theta)$ and finally to the $K$-dimensional output space $\mathcal{Y}$. Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}_i$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})$ where $\arg\max_y{\mathbf{y}^{+}}=y^+$ and $y^+$ denotes the index of the target class.

@@

\begin{enumerate}
    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
    \item (Actionability) Permutations $\Delta$ are subject to some actionability constraints.
    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is defined as above.
\end{enumerate}
where $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior distribution over inputs. For simplicity, we refer to a model with high explanatory capacity as \textbf{explainable}. 
\end{definition}

\noindent The characterization of faithfulness and plausibility in Def.
\ref{def-explainability} is the same as in \citet{altmeyer2024faithful},
with adapted notation. Intuitively, plausible counterfactuals are
consistent with the data and faithful counterfactuals are consistent
with what the model has learned about input data. Actionability
constraints in Def. \ref{def-explainability} vary and depend on the
context in which \(\mathbf{M}_\theta\) is deployed. In this work, we
focus on domain and mutability constraints for individual features
\(x_d\) for \(d=1,...,D\). We limit ourselves to classification tasks
for reasons discussed in Section~\ref{sec-discussion}.

@@

To train models with high explainability as defined in Def.
\ref{def-explainability}, we propose to leverage counterfactuals in the
following objective:

\begin{equation}\phantomsection\label{eq-obj}{
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) &+ \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{split}
}\end{equation} where \(\text{yloss}(\cdot)\) is a classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms are explained in detail below. For now, they can
be summarized as inducing explainability directly and indirectly by
penalizing: (1) the contrastive divergence, \(\text{div}(\cdot)\),
between mature counterfactuals \(\mathbf{x}_T^\prime\) and observed
samples \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the
target class \(y^+\), and, (2) the adversarial loss,
\(\text{advloss}(.)\), with respect to nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\). Finally, \(\text{ridge}(\cdot)\)
denotes a Ridge penalty (\(\ell_2\)-norm) that regularizes the magnitude
of the energy terms involved in \(\text{div}(\cdot)\)
\citep{du2019implicit}. The trade-off between the components can
governed through penalties \(\lambda_{\text{div}}\),
\(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\).

@@

As observed by \citet{grathwohl2020your}, any classifier can be
re-interpreted as a joint energy-based model (JEM) that learns to
discriminate output classes conditional on the observed (training)
samples from \(p(\mathbf{x})\) and the generated samples from
\(p_\theta(\mathbf{x})\). The authors show that JEMs can be trained to
perform well at both tasks by directly maximizing the joint
log-likelihood factorized as
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
The first term can be optimized using conventional cross-entropy as in
Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\),
\citet{grathwohl2020your} minimizes the contrastive divergence between
the observed samples from \(p(\mathbf{x})\) and generated samples from
\(p_\theta(\mathbf{x})\).

A key empirical finding in \citet{altmeyer2024faithful} was that JEMs
tend to do well with respect to the plausibility objective in Def.
\ref{def-explainability}. This follows directly if we consider samples
drawn from \(p_\theta(\mathbf{x})\) as counterfactuals because the JEM
objective effectively minimizes the divergence between the conditional
posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To generate samples,
\citet{grathwohl2020your} relies on Stochastic Gradient Langevin
Dynamics (SGLD) using an uninformative prior for initialization but we
depart from their methodology. Instead of SGLD, we propose to use
counterfactual explainers to generate counterfactuals of observed
training samples. Specifically, we have:

@@

The departure from SGLD of \citep{grathwohl2020your} allows us to tap
into the vast repertoire of explainers that have been proposed in the
literature to meet different desiderata. For example, many methods
facilitate the imposition of domain and mutability constraints. In
principle, any existing approach for generating counterfactual
explanations is viable, so long as it does not violate the faithfulness
condition. Like JEMs \citep{murphy2022probabilistic}, CT can be
considered a form of contrastive representation learning.

@@

Based on our analysis in Section~\ref{sec-lit}, counterfactuals
\(\mathbf{x}^\prime\) can be repurposed as additional training samples
\citep{balashankar2023improving, luu2023counterfactual} or adversarial
examples \citep{freiesleben2022intriguing, pawelczyk2022exploring}. This
leaves some flexibility wrt. the choice for \(\text{advloss}(\cdot)\) in
Equation~\ref{eq-obj}. An intuitive functional form, but likely not the
only sensible choice, is inspired by adversarial training:

\begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
}\end{equation} Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
perturbation to any single feature is at most \(\varepsilon\). This is
closely aligned with \citet{szegedy2013intriguing} that defines an
adversarial attack as an ``imperceptible non-random perturbation''.
Thus, we choose to work with a different distinction between CE and AE
than \citet{freiesleben2022intriguing} that considers misclassification
as the key distinguishing feature of AE. One of the key observations of
this work is that we can leverage CEs during training and get
adversarial examples essentially for free, which can be used to reap the
aforementioned benefits of adversarial training.

@@

Many existing counterfactual explainers support domain and mutability
constraints out-of-the-box. In fact, both types of constraints can be
implemented for any counterfactual explainer that relies on gradient
descent in the feature space for optimization
\citep{altmeyer2023explaining}. In this context, domain constraints can
be imposed by simply projecting counterfactuals back to the specified
domain, if the previous gradient step resulted in updated feature values
that were out-of-domain. Mutability constraints can similarly be
enforced by setting partial derivatives to zero to ensure that features
are only perturbed in the allowed direction, if at all.

Since such actionability constraints are binding at test time, we should
also impose them when generating \(\mathbf{x}^\prime\) during each
training iteration to inform model representations. Through their effect
on \(\mathbf{x}^\prime\), both types of constraints influence model
outcomes via Equation~\ref{eq-div}. Here it is crucial that we avoid
penalizing implausibility that arises due to mutability constraints. For
any mutability-constrained feature \(d\) this can be achieved by
enforcing \(\mathbf{x}^+[d] - \mathbf{x}^\prime[d]:=0\) whenever
perturbing \(\mathbf{x}^\prime[d]\) in the direction of
\(\mathbf{x}^+[d]\) would violate mutability constraints. Specifically,
we set \(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) if:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Feature \(d\) is strictly immutable in practice.
\item
  We have \(\mathbf{x}^+[d]>\mathbf{x}^\prime[d]\), but feature \(d\)
  can only be decreased in practice.
\item
  We have \(\mathbf{x}^+[d]<\mathbf{x}^\prime[d]\), but feature \(d\)
  can only be increased in practice.
\end{enumerate}

\noindent From a Bayesian perspective, setting
\(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) can be understood as
assuming a point mass prior for \(p(\mathbf{x}^+)\) with respect to
feature \(d\). Intuitively, we think of this simply in terms ignoring
implausibility costs with respect to immutable features, which
effectively forces the model to instead seek plausibility with respect
to the remaining features. This in turn results in lower overall
sensitivity to immutable features, which we demonstrate empirically for
different classifiers in Section~\ref{sec-experiments}. Under certain
conditions, this result holds theoretically:\footnote{For the proof, see
  the supplementary appendix.}

\begin{proposition}[Protecting Immutable Features]
\label{prp-mtblty}
Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$. If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is discriminative and mutable.
\end{proposition}

\noindent It is worth highlighting that
\mbox{Proposition \ref{prp-mtblty}} assumes independence of features.
This raises a valid concern about the effect of protecting immutable
features in the presence of proxies that remain unprotected. We address
this in Section~\ref{sec-discussion}.

@@

Suppose we are interested in predicting the likelihood that loan
applicants default on their credit. We have access to historical data on
previous loan takers comprised of a binary outcome variable
(\(y\in\{1=\text{default},2=\text{no default}\}\)) with two input
features: (1) the subjects' \emph{age}, which we define as immutable,
and (2) the subjects' existing level of \emph{debt}, which we define as
mutable.

We have simulated this scenario using synthetic data with independent
\emph{age} and \emph{debt} features, and Gaussian class-conditional
densities in Figure~\ref{fig-poc}. The four panels show the outcomes for
different training procedures using the same model architectures (a
linear classifier). In panels (a) and (c) we have trained the models
conventionally, while in panels (b) and (d) we used CT.

In all cases, all counterfactuals (stars) are valid---they have crossed
the decision boundary (green)---but their quality differs. In panel (a),
they are not plausible: they do not comply with the distribution of the
factuals in \(y^+\) to the point where they form a clearly
distinguishable cluster. In panel (b), they are highly plausible,
meeting the first objective of Def. \ref{def-explainability}. In panel
(c), the CEs involve substantial reductions in \emph{debt} for younger
applicants. By comparison, counterfactual paths are shorter on average
in panel (d) where we have protected the immutable \emph{age} as
described in Section~\ref{sec-constraints}. Due to the classifier's
lower sensitivity to \emph{age}, recommendations with respect to
\emph{debt} are much more homogenous and do not unfairly punish younger
individuals. These counterfactuals are also plausible with respect to
the mutable feature. Thus, we consider the model in panel (d) as the
most explainable according to Def. \ref{def-explainability}.

@@

\caption{\label{fig-poc}Illustration of how CT improves model
explainability: (a) conventional training, all mutable; (b) CT, all
mutable; (c) conventional, \emph{age} immutable; (d) CT, \emph{age}
immutable. The linear decision boundary is shown in green along with
training data colored according to their ground-truth label: \(y^-=1\)
in blue and \(y^+=2\) in orange. Stars indicate counterfactuals in the
target class.}

>>>>

This section introduces the Counterfactual Training objective. CT
combines ideas from adversarial training, counterfactual explanations,
and energy-based modelling with the explicit goal of producing models
whose learned representations align with plausible explanations that
further comply with user-defined actionability constraints.

In the context of counterfactual explanations, plausibility has broadly
been defined as the degree to which generated CEs comply with the
underlying data-generating process
\citep{altmeyer2024faithful, guidotti2022counterfactual, poyiadzi2020face}.
Plausibility is a necessary but insufficient condition for using CEs to
provide algorithmic recourse (AR) to individuals (negatively) affected
by opaque models. An AR recommendations must also be actionable, i.e.,
possible to attain by the recipient. A plausible CE for a rejected
20-year-old loan applicant, for example, might reveal that their
application would have been accepted, if only they had been 20 years
older. Ignoring all other features, this would comply with the
definition of plausibility if 40-year-old individuals were in fact more
credit-worthy on average than young adults. But of course this CE does
not qualify for providing actionable recourse to the applicant since
\emph{age} is not a (directly) mutable feature. Counterfactual training
aims to improve model explainability by aligning models with
counterfactuals that meet both desiderata: plausibility and
actionability. Formally, we define explainability as follows:

\begin{definition}[Model Explainability]
\label{def-explainability}
Let $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denote a supervised classification model that maps from the $D$-dimensional input space $\mathcal{X}$ to representations $\phi(\mathbf{x};\theta)$ and finally to the $K$-dimensional output space $\mathcal{Y}$. Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}_i$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})$, where $\arg\max_y{\mathbf{y}^{+}}=y^+$ is the index of the target class.

@@

\begin{enumerate}
    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$, where $A$ is some arbitrarily small region around $\mathbf{x}^{\prime}$.
    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$; $A$ as specified above.
    \item (Actionability) Permutations $\Delta$ are subject to some actionability constraints.
\end{enumerate}
and $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior distribution over inputs. For simplicity, we refer to a model with high explanatory capacity as \textbf{explainable} in this manuscript. 
\end{definition}

\noindent The characterization of faithfulness and plausibility in Def.
\ref{def-explainability} is the same as in \citet{altmeyer2024faithful},
with adapted notation. Intuitively, plausible counterfactuals are
consistent with the data and faithful counterfactuals are consistent
with what the model has learned about the input data. Actionability
constraints in Def. \ref{def-explainability} vary and depend on the
context in which \(\mathbf{M}_\theta\) is deployed. In this work, we
choose to only consider domain and mutability constraints for individual
features \(x_d\) for \(d=1,...,D\). We also limit ourselves to
classification tasks for reasons discussed in
Section~\ref{sec-discussion}.

@@

To train explainable models as defined in Def. \ref{def-explainability},
we propose to leverage counterfactuals in the following objective:

\begin{equation}\phantomsection\label{eq-obj}{
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) &+ \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{split}
}\end{equation} where \(\text{yloss}(\cdot)\) is a classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms are explained in detail below. For now, they can
be summarized as inducing explainability directly and indirectly by
penalizing (in 2\textsuperscript{nd} term) the contrastive divergence,
\(\text{div}(\cdot)\), between mature counterfactuals
\(\mathbf{x}_T^\prime\) and observed samples
\(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the target
class \(y^+\), and (in 3\textsuperscript{rd} term) the adversarial loss,
\(\text{advloss}(.)\), with respect to nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\). Finally, \(\text{ridge}(\cdot)\)
denotes a Ridge penalty (\(\ell_2\)-norm) that regularizes the magnitude
of the energy terms involved in \(\text{div}(\cdot)\)
\citep{du2019implicit}. The trade-off between the components can
governed through penalties \(\lambda_{\text{div}}\),
\(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\).

@@

\citet{grathwohl2020your} observed that any classifier can be
re-interpreted as a joint energy-based model that learns to discriminate
output classes conditional on the observed (training) samples from
\(p(\mathbf{x})\) and the generated samples from
\(p_\theta(\mathbf{x})\). \citet{grathwohl2020your} show that JEMs can
be trained to perform well at both tasks by directly maximizing the
joint log-likelihood:
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\),
where the first term can be optimized using cross-entropy as in
Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\), they
minimize the contrastive divergence between the observed samples from
\(p(\mathbf{x})\) and the generated samples from
\(p_\theta(\mathbf{x})\).

A key empirical finding of \citet{altmeyer2024faithful} was that JEMs
perform well on the plausibility objective in Def.
\ref{def-explainability}. This follows directly if we consider samples
drawn from \(p_\theta(\mathbf{x})\) as counterfactuals --- the JEM
objective effectively minimizes the divergence between the conditional
posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To generate samples,
\citet{grathwohl2020your} use Stochastic Gradient Langevin Dynamics
(SGLD) with an uninformative prior for initialization but we depart from
their methodology. Instead we propose to leverage counterfactual
explainers to generate counterfactuals of observed training samples.
Specifically, we have:

@@

The departure from SGLD allows us to tap into the vast repertoire of
explainers that have been proposed in the literature to meet different
desiderata. For example, many methods support domain and mutability
constraints. In principle, any existing approach for generating CEs is
viable, so long as it does not violate the faithfulness condition. Like
JEMs \citep{murphy2022probabilistic}, Counterfactual Training can be
considered a form of contrastive representation learning.

@@

Based on our analysis in Section~\ref{sec-lit}, counterfactuals
\(\mathbf{x}^\prime\) can be repurposed as additional training samples
\citep{balashankar2023improving, luu2023counterfactual} or adversarial
examples \citep{freiesleben2022intriguing, pawelczyk2022exploring}. This
leaves some flexibility with regards to the choice for the
\(\text{advloss}(\cdot)\) term in Equation~\ref{eq-obj}. An intuitive
functional form, but likely not the only sensible choice, is inspired by
adversarial training:

\begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
}\end{equation} Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
perturbation to any single feature is at most \(\varepsilon\). This is
closely aligned with \citet{szegedy2013intriguing} who define an
adversarial attack as an ``imperceptible non-random perturbation''.
Thus, we choose to work with a different distinction between CE and AE
than \citet{freiesleben2022intriguing} who consider misclassification as
the key distinguishing feature of AE. One of the key observations of
this work is that we can leverage CEs during training and get
adversarial examples essentially for free, which can be used to reap the
aforementioned benefits of adversarial training.

@@

Many existing counterfactual explainers support domain and mutability
constraints out-of-the-box. In fact, both types of constraints can be
implemented for any explainer that relies on gradient descent in the
feature space for optimization \citep{altmeyer2023explaining}. In this
context, domain constraints can be imposed by simply projecting
counterfactuals back to the specified domain, if the previous gradient
step resulted in updated feature values that were out-of-domain.
Similarly, mutability constraints can be enforced by setting partial
derivatives to zero to ensure that features are only perturbed in the
allowed direction, if at all.

Since actionability constraints are binding at test time, we should also
impose them when generating \(\mathbf{x}^\prime\) during each training
iteration to inform model representations. Through their effect on
\(\mathbf{x}^\prime\), both types of constraints influence model
outcomes via Equation~\ref{eq-div}. Here it is crucial that we avoid
penalizing implausibility that arises due to mutability constraints. For
any mutability-constrained feature \(d\) this can be achieved by
enforcing \(\mathbf{x}^+[d] - \mathbf{x}^\prime[d]:=0\) whenever
perturbing \(\mathbf{x}^\prime[d]\) in the direction of
\(\mathbf{x}^+[d]\) would violate mutability constraints. Specifically,
we set \(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) if:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Feature \(d\) is strictly immutable in practice.
\item
  \(\mathbf{x}^+[d]>\mathbf{x}^\prime[d]\), but \(d\) can only be
  decreased in practice.
\item
  \(\mathbf{x}^+[d]<\mathbf{x}^\prime[d]\), but \(d\) can only be
  increased in practice.
\end{enumerate}

\noindent From a Bayesian perspective, setting
\(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) can be understood as
assuming a point mass prior for \(p(\mathbf{x}^+)\) with respect to
feature \(d\). Intuitively, we think of this as ignoring implausibility
costs with respect to immutable features, which effectively forces the
model to instead seek plausibility with respect to the remaining
features. This in turn results in lower overall sensitivity to immutable
features, which we demonstrate empirically for different classifiers in
Section~\ref{sec-experiments}. Under certain conditions, this result
holds theoretically:\footnote{For the proof, see the supplementary
  appendix.}

\begin{proposition}[Protecting Immutable Features]
\label{prp-mtblty}
Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$. Assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is discriminative and mutable.
\end{proposition}

\noindent It is worth noting that \mbox{Proposition \ref{prp-mtblty}}
assumes independence of features. This raises a valid concern about the
effect of protecting immutable features in the presence of proxies that
remain unprotected. We address this in Section~\ref{sec-discussion}.

@@

Suppose we are interested in predicting the likelihood that loan
applicants default on their credit. We have access to historical data on
previous loan takers comprised of a binary outcome variable
(\(y\in\{1=\text{default},2=\text{no default}\}\)) with two input
features: (1) the subjects' \emph{age}, which we define as immutable,
and (2) the subjects' existing level of \emph{debt}, which we define as
mutable. We have simulated this scenario using synthetic data with
independent \emph{age} and \emph{debt} features, and Gaussian
class-conditional densities in Figure~\ref{fig-poc}. The four panels
show the outcomes for different training procedures using the same model
architecture (a linear classifier). In panels (a) and (c) we have
trained the models conventionally, while in panels (b) and (d) we have
applied CT.

In all four cases, all counterfactuals (stars) are valid---they have
crossed the decision boundary (green) into the target class---but their
quality differs. In panel (a), they are not plausible: they do not
comply with the distribution of the factuals in \(y^+\) to the point
where they form a clearly discernible cluster. In panel (b), they are
highly plausible, meeting the first objective of Def.
\ref{def-explainability}. In panel (c), the CEs involve substantial
reductions in \emph{debt} for younger applicants. By comparison,
counterfactual paths are shorter on average in panel (d) where we have
protected the immutable \emph{age} as described in
Section~\ref{sec-constraints}. Due to the classifier's lower sensitivity
to \emph{age}, recommendations with respect to \emph{debt} are much more
homogenous and do not unfairly punish younger individuals. These
counterfactuals are also plausible with respect to the mutable feature.
Thus, we consider the model in panel (d) as the most explainable
according to Def. \ref{def-explainability}.

@@

\caption{\label{fig-poc}Illustration of how CT improves model
explainability: (a) conventional training, all mutable; (b) CT, all
mutable; (c) conventional, \emph{age} immutable; (d) CT, \emph{age}
immutable. The linear decision boundary is shown in green along with
training data colored according to ground-truth labels: \(y^-=1\) (blue)
and \(y^+=2\) (orange). Stars indicate counterfactuals in the target
class.}
